Left-linear rules in adhesive categories

### 2024-02-13 icalp/fusione.tex

Technically, to acquire the expressiveness needed for capturing
fusions, i.e. the merge of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, to the best of our knowledge the
phenomena arising when devising a theory of indepedence and
concurrency for rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories
%(which capture most ``applications'') 
have not been systematically
explored~\cite{Ehrig1976,EHP:BRfTToHLRS,baldan2011adhesivity},
even if an investigation in the alternative setting of quasi-topoi has 
been presented in~\cite{BehrHK23}.

Here we perform this exploration working at the level of $\mathcal{M}$-adhesive categories, 
%i.e.such that adhesivity holds only for a class $\mathcal{M}$ of arrows, 
which encompass a large family of DPO-based formalisms adopted by pratictioners~\cite{xxx}.

>>>>

Technically, to acquire the expressiveness needed for capturing
fusions, i.e. the merge of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, to the best of our knowledge the
phenomena arising when devising a theory of independence and
concurrency for rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories
%(which capture most ``applications'') 
have not been systematically
explored~\cite{Ehrig1976,EHP:BRfTToHLRS,baldan2011adhesivity},
even if an investigation in the setting of quasi-topoi, considering general
non-linear rules yet restricting their applicability,
% of rules and focussing as far as DPO in concerned on quasi-adhesivity, 
has been presented in~\cite{BehrHK23}.

Here we perform this exploration at the level of $\mathcal{M}$-adhesive categories, 
which encompass a large family of DPO-based formalisms adopted by researchers and 
pratictioners alike~\cite{xxx}.
